%---------------------------------------------------------------------------%
%->> Frontmatter
%---------------------------------------------------------------------------%
%-
%-> 生成封面
%-
\maketitle% 生成中文封面
% \MAKETITLE% 生成英文封面
%-
%-> 作者声明
%-
% \makedeclaration% 生成声明页
%-
%-> 中文摘要
%-
% \intobmk\chapter*{摘\quad 要}% 显示在书签但不显示在目录
\chapter*{摘\quad 要}% 显示在书签但不显示在目录
\setcounter{page}{1}% 开始页码
\pagenumbering{Roman}% 页码符号
\addcontentsline{toc}{chapter}{摘\quad 要}

    为了研究形变核反应体系光学势随能量变化的关系，本工作首次测量了52-72~MeV六个能量点下$^{18}\mathrm{O}$+$^{152}\mathrm{Sm}$体系的准弹性和弹性散射角分布。弹性散射角分布由高动量分辨的Q3D磁谱仪测量得到，而准弹性散射（包括基态、第一和第二激发态等）角分布由反应靶室内的硅PIN阵列测量得到。在弹性散射角分布中观察到了显著的库仑虹抑制现象，表明该体系存在大形变导致的强库仑激发的耦合，导致弹性散射反应道被吸收而异常地快速下降。
    基于实验角分布数据，采用耦合道 ECIS95 程序提取了各能点下光学势的实部和虚部势深度，获得了其能量依赖行为。结果显示，随着能量降低，光学势虚部出现异常增强，且实、虚部之间不满足色散关系，表现出典型的反常阈异常特征。这一现象首次在形变核体系中被发现，表明除破裂效应外，强库仑激发等强吸收机制同样有可能导致光学势的反常阈行为。
    本研究为探索反常阈异常的起因提供了新的线索。

\keywords{光学模型；大形变核；准弹性散射；库仑虹抑制；阈异常; 反常阈异常}% 中文关键词
%-
%-> 英文摘要
%-
% \intobmk\chapter*{ABSTRACT}% 显示在书签但不显示在目录
\chapter*{ABSTRACT}% 显示在书签但不显示在目录
\addcontentsline{toc}{chapter}{ABSTRACT}
\linespread{1.25}\selectfont% 正文格式
To investigate the energy dependence of the optical potential in deformed nuclear reaction systems, this work reports the first measurement of quasi-elastic and elastic scattering angular distributions for the $^{18}\mathrm{O}$+$^{152}\mathrm{Sm}$Sm system at six incident energies ranging from 52 to 72 MeV. The elastic scattering angular distributions were measured using the Q3D magnetic spectrometer with high momentum resolution, while the quasi-elastic scattering angular distributions (including the ground state, first and second excited states) were obtained using a silicon detector array inside the reaction chamber. A significant suppression of the Coulomb rainbow was observed in the elastic scattering angular distributions, indicating the presence of strong Coulomb excitation coupling induced by the large deformation of the target nucleus, which leads to an anomalously rapid decrease in the elastic scattering channel due to enhanced absorption.
    Based on the experimental angular distribution data, the real and imaginary parts of the optical potential at each energy point were extracted using the coupled-channels code ECIS95, and their energy-dependent behavior was obtained. The results show that as the incident energy decreases, the imaginary part of the optical potential exhibits an anomalous enhancement, and the real and imaginary parts do not satisfy the dispersion relation, displaying typical characteristics of the anomalous threshold anomaly. This phenomenon is observed for the first time in a deformed nuclear system, indicating that in addition to breakup effects, strong absorption mechanisms such as intense Coulomb excitation can also lead to anomalous threshold behavior in the optical potential.
    This study provides new insights into the origin of the anomalous threshold anomaly.

\KEYWORDS{optical model; heavy deformed nuclei; quasi-elastic scattering; Coulomb rainbow suppression; threshold anomaly; abnormal threshold anomaly;}% 英文关键词
%---------------------------------------------------------------------------%
